Bucket-Sorted Independent Sets for Algebraic Multigrid

### 2007-09-18 bsis_paper/paper.tex
Edits and some new writing.
git-svn-id: file:///home/rblake/svn/pamg/trunk@24 c7f3fe13-8c30-0410-9127-9a9c42fe009c

\section{Coarse-Grid Selection}
\label{sec:cgs}
In AMG, coarse-grid selection is the process of selecting the degrees
of freedom for the coarse level problem. In classical AMG, degrees of
freedom on a coarse level are a subset of the degrees of freedom on a
finer level. Smoothed
aggregation~\cite{Vanek1996,Vanek2001,brezina2005} forms a
coarse-level degree of freedom by aggregating several fine-level
degrees of freedom. In this paper, we study coarsening for classical
AMG.

Given a set of fine-level degrees of freedom, $\Omega_k$, coarse-grid
selection for classical AMG selects a set of coarse-level degrees of
freedom $\Omega_{k+1} \subset \Omega_k$. Several approaches have been
developed for selecting the coarse grid.

Independent set-based algorithms are most numerous and all trace their
roots to the classical AMG coarsening algorithm, RS
coarsening~\cite{Ruge1987}. These methods are our focus and are
discussed in more detail in the next section.

Other methods take substantially different approaches to
coarsening. The coarsening technique in~\cite{Griebel2006} develops a
parallel coarsening algorithm that utilizes RS. On each processor, RS
selects several different coarse grids, and then one coarse grid is
selected per processor in a way to minimize special treatment on
processor boundaries. In~\cite{MacLachlan2006}, a greedy approach is
used to produce a good splitting of the unknowns into fine-grid and
coarse-grid sets. Subdomain blocking techniques~\cite{Krechel2001}
offer another approach for parallel coarse-grid selection by
decoupling coarse grids and alleviating the need for communication on
coarse levels. Compatible relaxation
approaches~\cite{Brandt2000,Livne2004,Brannick2005} form coarse grids
incrementally and decide when the coarse grid is sufficient through a
quality measure based on relaxation.

@@

Independent set-based coarse grid selection algorithms implement three
routines to select coarse grids: initialization, independent set
selection, and weight update. A general independent-set based
coarsening algorithm is shown in
Algorithm~\ref{alg:general-indep-set}.
\begin{lstlisting}[caption={Independent Set-Based Coarse Grid
Selection},label=alg:general-indep-set]
%\textsc{Indep-Set-Coarse-Grid-Selection}$(\Omega)$:%
$F \leftarrow \emptyset$
$C \leftarrow \emptyset$
$w \leftarrow\,$%\textsc{Initialize-Weights}$(\Omega)$%
while $\Omega \ne \emptyset$
   $D \leftarrow\,$%\textsc{Select-Indep-Set}$(w,\, \Omega)$%
   $(F,\, C,\, \Omega,\, w) \leftarrow\,$%\textsc{Update-Weights}$(F,\,
                                C,\, \Omega,\, w)$%
end
\end{lstlisting}

Each vertex in the graph is assigned a weight based on its merit to
becoming a $C$-point. The largest weights are conferred to vertices
best suited to become $C$-points. The quality of potential $C$-points
is quantified by \emph{strength of connection} measures. The classical
strength of connection measure is based entirely on the relative sizes
of off-diagonal nonzeros in $A$. In the classical definition, the set
of unknowns that unknown $i$ strongly depends upon is defined as
\begin{equation}
\label{2:eqn:S_i}
S_i = \left\{j: i \ne j \textrm{ and }|a_{ij}| \ge \theta \max_{k \ne
i}|a_{ik}|\right\},
\end{equation}
where $a_{ij}$ is the nonzero in row $i$, column $j$ of matrix $A$ and
$0 < \theta \le 1$. The set of unknowns that $i$ strongly influences,
denoted $S_i^T$, is defined as the set of unknowns that strongly
depend on $i$: $S_i^T = \{j: i \in S_j\}$.

@@

In classical coarsening, the independent set $D$ consists of a single
vertex $i$ in each iteration, where $w_i \ge w_j,\, \forall j \in S_i
\cup S_i^T$. Forming independent sets with only a single vertex is
inefficient in parallel, so parallel independent set methods use
different initialization techniques leading to the ability to form
larger independent sets. In these algorithms vertex $i$ is eligible to
be added to the independent set $D$ if the following condition is
satisfied:
\begin{equation}
  \label{eq:D-selection}
  w_i > w_j,\, \forall j \in S_i \cup S_i^T.
\end{equation}
The research presented here focuses on the independent set criterion
in~(\ref{eq:D-selection}).

The weight update routines used in independent set-based coarsening
algorithms are varied but primarily take one of two forms: routines
that decrease the values of weights and routines that increase the
value of weights. In all cases, a relatively large weight confers
large value for having the corresponding vertex become a $C$-point.

@@

Theory is developed in this section to prove coarse-grid selection
invariance in general independent set-based algorithms. The algorithms
considered thus far select independent sets using
(\ref{eq:D-selection}), meaning vertex $i$ is eligible to be in $D$ if
its weight is larger than the weights of vertices in its neighborhood,
which is defined as follows.
\begin{defn}
\label{def:neighborhood}
The \emph{neighborhood} of vertex $i$, denoted $\mathcal{N}_i$, is the
union of the sets containing the strong influences of $i$ and the
strong dependencies of $i$: $\mathcal{N}_i = S_i \cup S_i^T$.
\end{defn}
Recall the neighborhood of $i$
(Definition~\ref{3:def:neighborhood}) is the set of vertices strongly
influenced by $i$ or strongly influencing $i$. Algorithms relying on
different and larger neighborhoods, such as \emph{distance-$d$
neighborhoods}, are conceivable.
\begin{defn}
The \emph{distance-$d$ neighborhood} of $i$, denoted
$\mathcal{N}_i^d$, is the set of vertices within $d$ hops of $i$ in
the symmetrized strength matrix, excluding $i$. That is,
$\mathcal{N}_i^d = \left[\bigcup_{j \in
\mathcal{N}_i^{d-1}}\left(\mathcal{N}_j \cup \{j\}\right)\cup
\mathcal{N}_i\right] \setminus \{i\}$, where $d > 0$ and
$\mathcal{N}_i^0 = \emptyset$.
\end{defn}

For a general independent set-based coarse-grid selection algorithm, a
vertex $i$ is eligible to be added to $D$ if
\begin{equation}
\label{5:eqn:general-selection}
w_i > w_j\;\textrm{for all } j \in \mathcal{N}_i^s,
\end{equation}
where $\mathcal{N}_i^s$ is the \emph{selection neighborhood} of
$i$. Although the distance-$d$ neighborhood is a sensible choice for
the selection neighborhood, this discussion is not limited to such
cases. It is assumed, however, the matrix formed by $\mathcal{N}_*^s$
(i.e., the selection sets for all vertices) is symmetric. This is
equivalent to stating if $j \in \mathcal{N}_i^s$, then $i \in
\mathcal{N}_j^s$ for all $i$ and $j$ in the vertex set. Furthermore,
this discussion assumes the weight of vertex $i$ is only decremented
following weight updates and is never modified due to the assignment
of a vertex $j \notin \mathcal{N}_i^s$ to the $C$-point set.

@@

Vertices in $D_0$ are destined to become $C$-points regardless of the
algorithm used to build coarse grids. In the independent set-based
algorithms discussed in Chapters~\ref{chap:CGS}
and~\ref{chap:color-based}, all $D_0$ vertices become $C$-points in
the first iteration. Any algorithm constructed, however, eventually
selects all $D_0$ vertices as $C$-points, as proven in the following
lemma.
\begin{lem}
\label{5:lemma:D0_invariant}
Given the same selection neighborhoods and update heuristics, all
algorithms select vertices in $D_0$ as $C$-points.
\end{lem}
\begin{proof}
The proof follows from Lemma~\ref{5:lemma:C-lock}. All vertices in
$D_0$ satisfy the selection condition~(\ref{5:eqn:general-selection}),
so the assignment of any other $C$-point has no effect on the weight
of a $D_0$ vertex.
\end{proof}

>>>>

\section{Coarse-Grid Selection}
\label{sec:cgs}
In AMG, coarse-grid selection is the process of selecting the degrees
of freedom for the coarse level problem. In classical AMG, degrees of
freedom on a coarse level are a subset of the degrees of freedom on a
finer level. Smoothed
aggregation~\cite{Vanek1996,Vanek2001,brezina2005}, on the other hand,
forms a coarse-level degree of freedom by aggregating several
fine-level degrees of freedom. In this paper, we study coarsening for
classical AMG.

Given a set of fine-level degrees of freedom, $\Omega_k$, coarse-grid
selection for classical AMG selects a set of coarse-level degrees of
freedom $\Omega_{k+1} \subset \Omega_k$. Several approaches have been
developed for selecting the coarse grid, with independent set-based
approaches the most numerous. Independent set-based algorithms trace
their roots to the classical AMG coarsening algorithm, RS
coarsening~\cite{Ruge1987}. These methods are our focus and are
discussed in more detail in the next section.

Other methods take substantially different approaches to
coarsening. The technique in~\cite{Griebel2006} develops a parallel
coarsening algorithm that utilizes RS. On each processor, RS selects
several different coarse grids, and then one coarse grid is selected
per processor in a way to minimize special treatment on processor
boundaries. In~\cite{MacLachlan2006}, a greedy approach is used to
produce a good splitting of the unknowns into fine-grid and
coarse-grid sets. Subdomain blocking techniques~\cite{Krechel2001}
offer another approach for parallel coarse-grid selection by
decoupling coarse grids and alleviating the need for communication on
coarse levels. Compatible relaxation
approaches~\cite{Brandt2000,Livne2004,Brannick2005} form coarse grids
incrementally and decide when the coarse grid is sufficient through a
quality measure based on relaxation.

@@

Independent set-based coarse grid selection algorithms implement three
routines to select coarse grids: initialization, independent set
selection, and weight update. A general independent-set based
coarsening algorithm is shown in
Algorithm~\ref{alg:general-indep-set}.
\begin{lstlisting}[caption={Independent Set-Based Coarse Grid
Selection},label=alg:general-indep-set]
%\textsc{Indep-Set-Coarse-Grid-Selection}$(\Omega)$:%
$F \leftarrow \emptyset$
$C \leftarrow \emptyset$
$w \leftarrow\,$%\textsc{Initialize-Weights}$(\Omega)$%
while $\Omega \ne \emptyset$
   $D \leftarrow\,$%\textsc{Select-Indep-Set}$(w,\, \Omega)$%
   $(F,\, C,\, \Omega,\, w) \leftarrow\,$%\textsc{Update-Weights}$(F,\,C,\, \Omega,\, w)$%
end
\end{lstlisting}

Each vertex in the graph is assigned a weight based on its merit to
becoming a $C$-point. The largest weights are conferred to vertices
best suited to become $C$-points, where the quality of potential
$C$-points is quantified by \emph{strength of connection}
measures. The classical strength of connection measure is based
entirely on the relative sizes of off-diagonal nonzeros in $A$. In the
classical definition, the set of unknowns that unknown $i$ strongly
depends upon is defined as
\begin{equation}
\label{2:eqn:S_i}
S_i = \left\{j: i \ne j \textrm{ and }|a_{ij}| \ge \theta \max_{k \ne
i}|a_{ik}|\right\},
\end{equation}
where $a_{ij}$ is the nonzero in row $i$, column $j$ of matrix $A$ and
$0 < \theta \le 1$. The set of unknowns that $i$ strongly influences,
denoted $S_i^T$, is defined as the set of unknowns that strongly
depend on $i$: $S_i^T = \{j: i \in S_j\}$.

@@

In classical coarsening, the independent set $D$ consists of a single
vertex $i$ in each iteration, where $w_i \ge w_j,\, \forall j \in S_i
\cup S_i^T$. Forming independent sets with only a single vertex is
inefficient in parallel, so parallel independent set methods use
different initialization techniques designed to enable the selection
of larger independent sets. In these algorithms, vertex $i$ is
eligible to be added to the independent set $D$ if the following
condition is satisfied:
\begin{equation}
  \label{eq:D-selection}
  w_i > w_j,\, \forall j \in S_i \cup S_i^T.
\end{equation}
The research presented here focuses primarily on the independent set
criterion in~(\ref{eq:D-selection}).

The weight update routines used in independent set-based coarsening
algorithms are varied but primarily take one of two forms: routines
that decrease the values of weights and routines that increase the
value of weights. In all cases, a vertex with a large weight has high
fitness for becoming a $C$-point.

@@

Theory is developed in this section to prove coarse-grid selection
invariance in general independent set-based algorithms. The algorithms
considered thus far select independent sets using
(\ref{eq:D-selection}), meaning vertex $i$ is eligible to be in $D$ if
its weight is larger than the weights of vertices in its neighborhood,
which is defined as follows.
\begin{defn}
\label{def:neighborhood}
The \emph{neighborhood} of vertex $i$, denoted $\mathcal{N}_i$, is the
union of the sets containing the strong influences of $i$ and the
strong dependencies of $i$: $\mathcal{N}_i = S_i \cup S_i^T$.
\end{defn}
Algorithms that rely on different and larger neighborhoods, such as
\emph{distance-$d$ neighborhoods}, are conceivable.
\begin{defn}
The \emph{distance-$d$ neighborhood} of $i$, denoted
$\mathcal{N}_i^d$, is the set of vertices within $d$ hops of $i$ in
the symmetrized strength matrix, excluding $i$. That is,
$\mathcal{N}_i^d = \left[\bigcup_{j \in
\mathcal{N}_i^{d-1}}\left(\mathcal{N}_j \cup \{j\}\right)\cup
\mathcal{N}_i\right] \setminus \{i\}$, where $d > 0$ and
$\mathcal{N}_i^0 = \emptyset$.
\end{defn}

For a general independent set-based coarse-grid selection algorithm, a
vertex $i$ is eligible to be added to $D$ if
\begin{equation}
\label{5:eqn:general-selection}
w_i > w_j\;\textrm{for all } j \in \mathcal{N}_i^s,
\end{equation}
where $\mathcal{N}_i^s$ is the \emph{selection neighborhood} of
$i$. Although the distance-$d$ neighborhood is a sensible choice for
the selection neighborhood, this discussion is not limited to such
cases. It is assumed, however, the matrix formed by $\mathcal{N}_*^s$
(i.e., the selection sets for all vertices) is symmetric. This is
equivalent to stating if $j \in \mathcal{N}_i^s$, then $i \in
\mathcal{N}_j^s$ for all $i$ and $j$ in the vertex set. Furthermore,
this discussion assumes the weight of vertex $i$ is only decremented
following weight updates and is unaffected by the assignment of a
vertex $j \notin \mathcal{N}_i^s$ to the $C$-point set.

@@

Vertices in $D_0$ are destined to become $C$-points regardless of the
algorithm used to build coarse grids. In most independent set-based
algorithms developed so far, all $D_0$ vertices become $C$-points in
the first iteration. Any algorithm constructed, however, eventually
selects all $D_0$ vertices as $C$-points, as proven in the following
lemma.
\begin{lem}
\label{5:lemma:D0_invariant}
Given the same selection neighborhoods and update heuristics, all
algorithms select vertices in $D_0$ as $C$-points.
\end{lem}
\begin{proof}
The proof follows from Lemma~\ref{5:lemma:C-lock}. All vertices in
$D_0$ satisfy the selection condition~(\ref{5:eqn:general-selection}),
so the assignment of any other $C$-point has no effect on the weight
of a $D_0$ vertex.
\end{proof}
